Dynamic Witnesses for Static Type Errors

### 2015-11-20 paper/pldi16-submission/overview.tex
tweak overview

Returning to @fac@, given a hole as input we will narrow the hole
to an @int@ upon reaching the @<=@ test.
%
At this point we choose a random @int@ for the instantiation and
concrete execution takes over entirely leading us to the expected crash
in the multiplication.

\paragraph{Witness Generality}
We show in \S~\ref{sec:soundness} our lazy instantiation of holes
produces \emph{general witnesses}.
%
That is, we show that if ``executing''
a function with a hole as input causes the
function to ``go wrong'', then there is
\emph{no possible} type for the function.
%
In other words, for \emph{any} type you might
assign to the function, there exist inhabitants
that will cause the function to go wrong.

\paragraph{Problem: How many inputs does a function take?}
%
There is another wrinkle though, how did we know
that @fac@ takes a single argument instead of two
(or none)?
%
It is clear, syntactically, that @fac@ takes \emph{at least} one
argument, but in a higher-order language with currying syntax can be
deceiving.
%
Consider the following definition:
%
\begin{code}
  let incAllByOne = List.map (+ 1)
\end{code}
%
Is @incAllByOne@ a function?
%
If so, how many arguments does it take?
%
The \ocaml compiler deduces that @incAllByOne@ takes a single argument
because @List.map@'s \emph{type} says it takes two arguments, and it is
partially applied to @(+ 1)@.
%
As we are dealing with ill-typed programs we do not have the luxury of
typing information.

@@

\subsection{Visualizing Witnesses}
\label{sec:visual-witness}
We have shown how to reliably find witnesses to type errors in \ocaml,
but this does not fully address our original goal -- to \emph{explain}
the errors.
%
Having identified an input vector that triggers a bug, a common next
step is to step through the program with a \emph{debugger} to observe
how the program evolves.
%
The existing debuggers and interpreters for \ocaml assume a type-correct
program, so unfortunately we cannot use them off-the-shelf.
%
Instead we extend our search for witnesses to produce an execution
trace.

@@

\paragraph{Interacting with the graph}
The reduction graph is useful for formulating and executing traversals,
but displaying it all at once would quickly become overwhelming.
%
Our interaction begins by displaying a big-step reduction, \ie the
witness followed by the stuck term.
%
The user can then progressively fill in the hidden steps of the
computation by selecting a visible term and choosing one of the
applicable traversal strategies --- formalized in
\S~\ref{sec:explaining} --- to insert another term into the
visualization.

@@

@fac 0@ seems like the right thing to do so we choose to ``step into''
it, which inserts a new multi-step transition from @fac 0@ to @true@.
%
Finally, we take a ``step forward'' from @fac 0@, bringing us to @fac@'s
body. If we mouse over the body term we will see a popup with the
environment at this point, notably telling us that @n = 0@. At this point
it is clear that @fac@ handles the @n <= 0@ case incorrectly and should
instead return an @int@.

>>>>

Returning to @fac@, given a hole as input we will narrow the hole
to an @int@ upon reaching the @<=@ test.
%
At this point we choose a random @int@ for the instantiation and
concrete execution takes over entirely, leading us to the expected crash
in the multiplication.

\paragraph{Witness Generality}
We show in \S~\ref{sec:soundness} that our lazy instantiation of holes
produces \emph{general witnesses}.
%
That is, we show that if ``executing''
a function with a hole as input causes the
function to ``go wrong'', then there is
\emph{no possible} type for the function.
%
In other words, for \emph{any} type you might
assign to the function, there exist inhabitants
that will cause the function to go wrong.

\paragraph{Problem: How many inputs does a function take?}
%
There is another wrinkle though, how did we know
that @fac@ takes a single argument instead of two
(or none)?
%
It is clear, syntactically, that @fac@ takes \emph{at least} one
argument, but in a higher-order language with currying, syntax can be
deceiving.
%
Consider the following definition:
%
\begin{code}
  let incAllByOne = List.map (+ 1)
\end{code}
%
Is @incAllByOne@ a function?
%
If so, how many arguments does it take?
%
The \ocaml\ compiler deduces that @incAllByOne@ takes a single argument
because @List.map@'s \emph{type} says it takes two arguments, and it is
partially applied to @(+ 1)@.
%
As we are dealing with ill-typed programs we do not have the luxury of
typing information.

@@

\subsection{Visualizing Witnesses}
\label{sec:visual-witness}
We have shown how to reliably find witnesses to type errors in \ocaml,
but this does not fully address our original goal --- to \emph{explain}
the errors.
%
Having identified an input vector that triggers a bug, a common next
step is to step through the program with a \emph{debugger} to observe
how the program evolves.
%
The existing debuggers and interpreters for \ocaml\ assume a type-correct
program, so unfortunately we cannot use them off-the-shelf.
%
Instead we extend our search for witnesses to produce an execution
trace.

@@

\paragraph{Interacting with the graph}
The reduction graph is useful for formulating and executing traversals,
but displaying it all at once would quickly become overwhelming.
%
Our interaction begins by displaying a big-step reduction, \ie the
witness followed by the stuck term.
%
The user can then progressively fill in the hidden steps of the
computation by selecting a visible term and choosing one of the
applicable traversal strategies --- formalized in
\S~\ref{sec:interactive} --- to insert another term into the
visualization.

@@

@fac 0@ seems like the right thing to do so we choose to ``step into''
it, which inserts a new multi-step transition from @fac 0@ to @true@.
%
Finally, we take a ``step forward'' from \hbox{@fac 0@,} bringing us to @fac@'s
body. If we mouse over the body term we will see a popup with the
environment at this point, notably telling us that @n = 0@. At this point
it is clear that @fac@ handles the @n <= 0@ case incorrectly and should
instead return an @int@.
